Tropical Mathematics and the Lambda-Calculus

### 2023-05-15 TLLA23/BarbaPisto.tex
commit

\title{Tropical methods in Lambda-Calculus}
\author{Davide Barbarossa
\institute{DISI, Universit\`a di Bologna}
\email{davide.barbarossa@unibo.it}
\and
\qquad\qquad Paolo Pistone
\institute{\qquad\qquad\qquad DISI, Universit\`a di Bologna}
\email{\qquad\qquad\qquad paolo.pistone@unibo.it}
}
\def\titlerunning{Trends in Linear Logic and Applications}
\def\authorrunning{D.\ Barbarossa and P.\ Pistone}

@@

Two different quantitative approaches have received considerable attention from the programming language community in recent years: %(e.g.\ \cite{decarvalho2018, Accattoli2022}, \cite{Breuvart2018, PistoneLICS2022}, \cite{Barthe_2012} \cite{Reed2010, dallago}, \cite{Ehrhard2011, Staton2017}, \cite{difflambda}), on the one hand one there is 
the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk} is based on the fact that when considering e.g.\ probabilistic %or approximate 
computation, %rather than asking whether two programs compute \emph{the same} function,
it is natural to ask whether two programs compute functions which %do not differ \emph{too much}
approximate each other, instead of equality of functions.
This led to denotational frameworks in which types are endowed with a metric \cite{Reed2010},\cite{Bonchi2018}, \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
On the other hand, there is the approach based on \emph{differential} \cite{difflambda}, \cite{difflambda}, \cite{Manzo2013, Breuvart2018, PistoneLICS2022} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
This led to syntactic or denotational frameworks in which one can define a \emph{Taylor expansion} of programs.
In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
In metric semantics, linear programs correspond to \emph{non-expansive} maps, i.e.\ that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret programs with a fixed duplication bound as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs gives rise to \emph{power series}.

At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
in metric models $n$ times duplication results in a $n$-Lipschitz \emph{linear} function $n\cdot x$, while in differential models this results in a non-Lipschitz \emph{polynomial} function $x^{n}$.
Our main observation is that 
this gap is overcome once we interpret these functions in the framework of tropical mathematics where, for instance, $x^{n}$ precisely reads as $n\cdot x$.
Tropical mathematics \cite{Simon} is a well established algebraic and geometrical framework, with tight connections with optimisation theory \cite{Sturmfelds}, where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
For instance, the polynomial $p(x,y)=x^{2}+xy^{2}+y^{3}$, when interpreted over the tropical semiring, translates as the piecewise linear function
$
\trop f(\alpha,\beta)=\min\{2\alpha, \alpha+2\beta, 3\beta\}
$.

@@

@@

We let the \emph{tropical semiring}  $\Lawv$, the structure at the heart of tropical mathematics, be $[0,\infty]$ with addition $\min$ and multiplication $+$.
This coincides with the \emph{Lawvere quantale} $\Lawv$ \cite{Hofmann2014, Stubbe2014}, i.e.\ $[0,\infty]$ with order $\geq$ and usual $+$ as the monoid action. 
$\Lawv$ is at the heart of the categorical study of metric spaces initiated by Lawvere \cite{Lawvere1973}, and we will take this point of view in the last section.
A \emph{tropical polynomial} is a piece-wise linear function $\varphi:\Lawv\to \Lawv$ of the form $\varphi(x)=\min_{j=1,\dots,k}\{i_{j}x+\matr \varphi_{i_{j}}\}$, with $i_{j}\in\N$, $\matr \varphi_{i_{j}}\in\Lawv$ and $k$ finite.
Those are always Lipschitz functions.
For example, $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$, plotted in Fig~\ref{fig:plot1}.
A \emph{tropical root} of $\varphi$ is a point $x\in \Lawv$ where the minimum defining $\varphi$ is attained at least twice.%, i.e.\ where $\varphi$ is not differentiable, i.e.~where the slope of $\varphi$ changes.
E.g., the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, $i \leq n$.
%With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as $f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
A \emph{tropical Laurent series} (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function $\varphi:\Lawv\to \Lawv$ of the form $\varphi(x)=\inf_{n\in\N}\{nx+\matr \varphi_{n}\}$, with $\matr \varphi_{n}\in\Lawv$.
That is, a tLs is a ``limit'' of tropical polynomials of higher and higher degree.
For example $\varphi(x):=\inf_{i\in\N}\set{ix+2^{-i}}$ %that we will take as our running example, 
is the ``limit'' of the $\varphi_{n}$, see Fig~\ref{fig:plot1}. %Since $\inf$s are not in general $\min$s, the behavior of tLS may be less predictable than that of tropical polynomials.
%For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points.
Finally, for a polynomial/power series $f(x)=\sum_n a_nx^n$, one defines its \emph{tropicalization} $\trop f(\alpha):=\inf_n\set{-\log a_n+n\alpha}$.
TLs may be non Lipschitz, and their study is less developed than that of tropical polynomials.

@@

The study of matrices with values over the tropical semiring is a special case of the
\emph{weighted relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus and linear logic:
for a fixed \emph{continuous} semi-ring $Q$, take the category $\QREL$ whose objects are sets and $\HOM \QREL X Y=Q^{X\times Y}$ (set-indexed matrices with coefficients in $Q$). %(\cite{Manzo2013} would call it $Q^\Pi$).
As expected, $Q^X$ is a $Q$-semimodule and we can identify $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.
Taking $Q:=\Lawv$ we obtain the \emph{tropical weighted relational model} $\LREL$.
%, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
%The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring (w.r.t.\ its quantale order $\preceq$.
%This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
Remark that the composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
reads as \ $(s\circ t)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
similarly, linear maps $f:\Lawv^X\to \Lawv^Y$ are of shape $f(x)_b=\inf_{a\in X} \set{x_a+\matr f_{a,b}}$, for some matrix $\matr f\in Q^{X\times Y}$ and are precisely those induced by $\matr f\in\HOM \LREL X Y$. We call them \emph{tropical linear}.
%\end{remark}
By applying known results (taken from \cite{Manzo2013},  {\color{red}CITARE LEMAYYYY}), one obtains that $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus:
$\LREL$ is a SMCC, i.e.\ a model of the linear $\STLC$;
the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$, where $!$ is the usual multiset comonad (so $!X$ is the set of finite multisets on $X$).
Here, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ given by:
$
 \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
 \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
Furthermore, $!$ can be decomposed into a family of \emph{graded} exponentials $(!_n)_{n\in\N}$ turning $(\LREL,(!_n)_{n\in\N})$ in a model for $\BSTLC$. 
Finally, $\LREL_!$ can be equipped with a \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined by: $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise. %(where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
This turns $(\LREL_!,D)$ into a CC$\partial$C, i.e.\ a model of the differential $\lambda$-calculus.
Remember that the (quantitative) Taylor expansion of an ordinary $\lambda$-term $M$ is an inductively defined series $\Te{M}$ of differential $\lambda$-terms, the only non-trivial case being $\Te{PQ} :=  \sum_{k=0}^{\infty} (\Der^{k}[P] \cdot Q^{k})0$.
It can be seen that the morphisms of $\LREL_!$ can always be Taylor expanded, and the series interpreting in $\LREL_!$ the Taylor expansion of a $\STLC$-term $M$, converges to the interpretation of $M$.
The Taylor formula decomposes an unbounded application as a limit of bounded ones.
So one might well ask whether this formula interprets a $\lambda$-term 
as a limit of Lipschitz maps and thus bridging, in a sense, the metric and differential approaches previously mentioned.  
%Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
But terms with bounded applications correspond to \emph{polynomials}, usually non-Lipschitz functions. 
Yet, if these polynomials were tropical, the Taylor expansion would really decompose $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps: just like in mathematical analysis, unbounded application could be seen 
as a limit of \emph{more and more sensitive} operations.
This is precisely the content of our Corollary~\ref{cor:main}(3).

\section{Tropical Laurent Series}

 As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a \emph{linear} map $\Lawv^{!X}\to\Lawv^Y$, but we can also ``express it in the base $X$'' and see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting 
 $t^!(x):=t\circ_! x$.
%  (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
% This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
 Concretely, we have $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$
 where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
 These functions correspond then to generalised tLs, i.e.\ with possibly infinitely many variables (as many as the elements of $X$).
For $X=Y=\set{*}$, we get usual tLs of one variable.
Instead, if the support $\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ of $f$ is \emph{finite}, we get %maps of shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, i.e.\ 
tropical polynomials in possibly infinitely many variables.
In this section we study the properties of generalised tLs, from the point of view of mathematical analysis.
%In the following we will also refer to them as tLs.

@@

@@

\end{tikzpicture}
\caption{Plot of the tropical polynomials $\varphi_1,\varphi_2,\varphi_3,\varphi_4$ (from top to bottom), and of their limit tLs $\varphi$ (in violet).}
\label{fig:plot1}
%\end{figure} 
\end{wrapfigure} %%

\begin{proposition}\label{prop:nondecr+conc}
 Any tLs $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
\end{proposition}

%In Example~\ref{ex:famous_ex}, 
Look at Fig~\ref{fig:plot1}: it appears that $\varphi$ behaves \emph{locally} like the polynomials $\varphi_{n}$. 
%In particular, for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$. More, precisely, $\varphi$ coincides with $\varphi_{n}$ for $\epsilon \geq 2^{-(n+1)}$, the smallest tropical root of $\varphi_{n}$.
However, at
%
%$x\in [\epsilon,\infty]$, with $\epsilon>0$
%It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
 $x=0$ we have that $\varphi(x=0)=\inf_{i\in\N} 2^{-i}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
Also, while the derivative of $\varphi$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
This phenomenon is reminiscent of [Example 7, \cite{Ehrhard2005}].
%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019 which actually motivated our first investigations.
In fact, these properties are shared by all tLs with \emph{finitely} many variables, as shown by the following result (we identify $!\set{1,\dots,k}$ with $\N^k$, so the matrix of a tLs $f$ with variables $x=x_1,\dots,x_k$ can be given as a $\matr f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, where $nx$ is the scalar product).

>>>>

\title{Tropical Mathematics and the Lambda-Calculus}
\author{Davide Barbarossa
\institute{DISI, Universit\`a di Bologna}
\email{davide.barbarossa@unibo.it}
\and
\qquad\qquad Paolo Pistone
\institute{\qquad\qquad\qquad DISI, Universit\`a di Bologna}
\email{\qquad\qquad\qquad paolo.pistone2@unibo.it}
}
\def\titlerunning{Trends in Linear Logic and Applications}
\def\authorrunning{D.\ Barbarossa and P.\ Pistone}

@@

In recent years, more and more interest in the programming language community has been directed towards the study of quantitative properties of programs like computing the number of computation steps or convergence probabilities, as opposed to purely qualitative properties like termination or program equivalence. 
In particular, two different quantitative approaches have received considerable attention from the programming language community. On the one hand, the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk} is based on the observation that probabilistic or numerical algorithms are not though to compute a target function $f$ \emph{exactly}, but only in an approximate way; hence, one is usually interested to know whether the computed function is \emph{not too far} from $f$, rather than equal to $f$. This led to study denotational frameworks in which types are endowed with metrics measuring similarities in program behavior \cite{Reed2010},\cite{Bonchi2018}, \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
On the other hand, there is the approach based on \emph{differential} \cite{difflambda}, \cite{difflambda}, \cite{Manzo2013, Breuvart2018, PistoneLICS2022} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the \emph{relational semantics}\cite{Manzo2012, Manzo2013, dill} and \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}.
This led to study syntactic or denotational frameworks in which one can define a \emph{Taylor expansion} of programs.

In both approaches a crucial role is played by the notion of \emph{linearity}, in the sense of linear logic, i.e.~of using inputs exactly once. In metric semantics, linear programs correspond to \emph{non-expansive} functions, i.e.\ maps that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret programs with a fixed duplication bound as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs gives rise to \emph{power series}.

The starting observation of this work is that, at a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
in metric models $n$ times duplication results in a $n$-Lipschitz \emph{linear} function $n\cdot x$, while in differential models this results in a non-Lipschitz \emph{polynomial} function $x^{n}$.
 At the same time, 
this gap may be overcome once we interpret these functions in the framework of tropical mathematics where, for instance, $x^{n}$ precisely reads as $n\cdot x$.

Tropical mathematics \cite{Simon} is a well established algebraic and geometrical framework, with tight connections with optimisation theory \cite{Sturmfelds}, where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
For instance, the polynomial $p(x,y)=x^{2}+xy^{2}+y^{3}$, when interpreted over the tropical semiring, translates as the piecewise linear function
$
\trop f(\alpha,\beta)=\min\{2\alpha, \alpha+2\beta, 3\beta\}
$.

@@

\section{The Tropical Semantics of Linear Logic }

@@

We let the \emph{tropical semiring}  $\Lawv$, the structure at the heart of tropical mathematics, be $[0,\infty]$ with addition $\min$ and multiplication $+$.
This coincides with the \emph{Lawvere quantale} $\Lawv$ \cite{Hofmann2014, Stubbe2014}, i.e.\ $[0,\infty]$ with order $\geq$ and usual $+$ as the monoid action. 
$\Lawv$ is at the heart of the categorical study of metric spaces initiated by Lawvere \cite{Lawvere1973}, and we will take this point of view in the last section.
A \emph{tropical polynomial} is a piece-wise linear function $\varphi:\Lawv\to \Lawv$ of the form $\varphi(x)=\min_{j=1,\dots,k}\{i_{j}x+\matr \varphi_{i_{j}}\}$, with $i_{j}\in\N$, $\matr \varphi_{i_{j}}\in\Lawv$ and $k$ finite.
Those are always Lipschitz functions.
For example, $\varphi_{n}(x)=\min_{i\leq n}\{ix+2^{-i}\}$, plotted in Fig~\ref{fig:plot1}.
A \emph{tropical root} of $\varphi$ is a point $x\in \Lawv$ where the minimum defining $\varphi$ is attained at least twice. %, i.e.\ where $\varphi$ is not differentiable, i.e.~where the slope of $\varphi$ changes.
E.g., the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, $i \leq n$.
%With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as $f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
A \emph{tropical Laurent series} (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function $\varphi:\Lawv\to \Lawv$ of the form $\varphi(x)=\inf_{n\in\N}\{nx+\matr \varphi_{n}\}$, with $\matr \varphi_{n}\in\Lawv$.
That is, a tLs is a ``limit'' of tropical polynomials of higher and higher degree.
For example $\varphi(x):=\inf_{i\in\N}\set{ix+2^{-i}}$ %that we will take as our running example, 
is the ``limit'' of the $\varphi_{n}$, see Fig~\ref{fig:plot1}. %Since $\inf$s are not in general $\min$s, the behavior of tLS may be less predictable than that of tropical polynomials.
%For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points.
Finally, for a polynomial/power series $f(x)=\sum_n a_nx^n$, one defines its \emph{tropicalization} $\trop f(\alpha):=\inf_n\set{-\log a_n+n\alpha}$.
TLs are in general not Lipschitz, and their study is less developed than that of tropical polynomials.

@@

The study of matrices with values over the tropical semiring is a special case of the
\emph{weighted relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus and linear logic:
for a fixed \emph{continuous} semi-ring $Q$, take the category $\QREL$ whose objects are sets and $\HOM \QREL X Y=Q^{X\times Y}$ (set-indexed matrices with coefficients in $Q$). %(\cite{Manzo2013} would call it $Q^\Pi$).
As expected, $Q^X$ is a $Q$-semimodule and we can identify $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.
Taking $Q:=\Lawv$ we obtain the \emph{tropical weighted relational model} $\LREL$.
%, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
%The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring (w.r.t.\ its quantale order $\preceq$.
%This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
Remark that the composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
reads as \ $(s\circ t)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
similarly, linear maps $f:\Lawv^X\to \Lawv^Y$ are of shape $f(x)_b=\inf_{a\in X} \set{x_a+\matr f_{a,b}}$, for some matrix $\matr f\in Q^{X\times Y}$ and are precisely those induced by $\matr f\in\HOM \LREL X Y$. We call them \emph{tropical linear}.
%\end{remark}
By applying known results (taken from \cite{Manzo2013},  {\color{red}CITARE LEMAYYYY}), one obtains that $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus:
$\LREL$ is a SMCC, i.e.\ a model of the linear $\STLC$;
the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$, where $!$ is the usual multiset comonad (so $!X$ is the set of finite multisets on $X$).
Here, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ where $(s\circ_! t)_{\mu,c}$ given by:
$
 \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
 \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
 As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a \emph{linear} map $\Lawv^{!X}\to\Lawv^Y$, but we can also ``express it in the base $X$'' and see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting 
 $t^!(x):=t\circ_! x$.
%  (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
% This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
 Concretely, we have $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$
 where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
 These functions correspond then to generalised tLs, i.e.\ with possibly infinitely many variables (as many as the elements of $X$).
For $X=Y=\set{*}$, we get usual tLs of one variable.
Instead, if the support $\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ of $f$ is \emph{finite}, we get %maps of shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, i.e.\ 
tropical polynomials in possibly infinitely many variables.
Furthermore, $!$ can be decomposed into a family of \emph{graded} exponentials $(!_n)_{n\in\N}$ turning $(\LREL,(!_n)_{n\in\N})$ in a model for a $\mathit{BLL}$-style simply typed $\lambda$-calculus, that we call $\BSTLC$ (see e.g.~\cite{Brunel2014}).

Finally, $\LREL_!$ can be equipped with a \emph{tropical differential operator} $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined by: $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise. %(where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
This turns $(\LREL_!,D)$ into a CC$\partial$C, i.e.\ a model of the differential $\lambda$-calculus.
Remember that the (quantitative) Taylor expansion of an ordinary $\lambda$-term $M$ is an inductively defined series $\Te{M}$ of differential $\lambda$-terms, the only non-trivial case being $\Te{PQ} :=  \sum_{k=0}^{\infty} (\Der^{k}[P] \cdot Q^{k})0$.
It can be seen that the morphisms of $\LREL_!$ can always be Taylor expanded, and the series interpreting in $\LREL_!$ the Taylor expansion of a $\STLC$-term $M$, converges to the interpretation of $M$.

%
%
%The Taylor formula decomposes an unbounded application as a limit of bounded ones.
%So one might well ask whether this formula interprets a $\lambda$-term 
%as a limit of Lipschitz maps and thus bridging, in a sense, the metric and differential approaches previously mentioned.  
%%Here, a natural direction to look for is the \emph{relational semantics}, i.e.~the somehow canonical ``Taylor'' semantics for $\STDLC$. 
%But terms with bounded applications correspond to \emph{polynomials}, usually non-Lipschitz functions. 
%Yet, if these polynomials were tropical, the Taylor expansion would really decompose $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps: just like in mathematical analysis, unbounded application could be seen 
%as a limit of \emph{more and more sensitive} operations.
%This is precisely the content of our Corollary~\ref{cor:main}(3).

\section{Tropical Laurent Series Yield a Metric Semantics of $\STLC$}

The main goal of this section is to show that the interpretation of $\STLC$ in the tropical relational model yields a metric semantics, where the spaces $\Lawv^{X}$ are endowed with the $\|\cdot\|_{\infty}$-norm metric, and programs are interpreted as \emph{locally Lipschitz} maps:
\begin{theorem}\label{cor:main}
For any $\lambda$-term $M$:
\begin{varenumerate}
\item if $\Gamma \vdash_{\BSTLC} M:A$, then $\model M^!:\model\Gamma \to \model A$ is a \emph{Lipschitz} map.
\item if $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is a \emph{locally Lipschitz} map. Moreover, the Taylor expansion decomposes $\model M^!$ into an $\inf$ of \emph{Lipschitz}~maps.
\end{varenumerate}
\end{theorem}
Recall that the Taylor formula decomposes an unbounded application as a limit of bounded ones; the result above presents this as a decomposition of an higher-order program as a limit of Lipschitz maps, and provides thus a bridge between the metric and the differential approaches.

The proof of the result above passes through the study of topological and metric properties of tLs. 
We first studied tLs with respect to the topology induced by the $\|\cdot\|_{\infty}$-norm.
%
%In this section we study the properties of generalised tLs, from the point of view of mathematical analysis.
%In the following we will also refer to them as tLs.

@@

\begin{proposition}\label{prop:nondecr+conc}
 Any tLs $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order, and continuous w.r.t.~the norm $\|\cdot\|_{\infty}$.
\end{proposition}

%
%The norm $\norm{\cdot}_\infty$ on $\BB R$ naturally induces a metric $\norm{x-y}_{\infty}$ over the spaces $\Lawv^{X}$.
%We will show that tLs satisfy suitable Lipschitz properties  w.r.t.~these metrics.
%\emph{Wlog}, we will often take $Y=\set{*}$.

Let us now consider metric properties. First, it can be seen that all tropical \emph{linear} functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive. 
%\begin{proof}[Proof sketch]
%Using the fact that $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
%the problem reduces to checking that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
This result shows that, in analogy with that happens in usual metric semantics, linear programs are interpreted by non-expansive functions. 
%\begin{proof}
%Using $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
%first observe that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$; we now have
%$|f(\B x)_{b}-f(\B y)_{b}| \leq |(\inf_{a\in X}\matr f_{a,b}-\B x_{a})-(\inf_{a\in X}\matr f_{a,b}-\B y_{a})| \leq
%\sup_{a\in X}|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})|\leq 
% \| \B x-\B y\|_{\infty}$.
%\end{proof}
More generally, linear maps with bounded exponentials yield Lipschitz maps:
\begin{proposition}\label{prop:boundedlip}
If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a bounded matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
\end{proposition}
%\begin{proof}[Proof sketch]
%It follows from the above lines about tropical linear functions, plus the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $! x$ to finite multisets of degree $\leq n$.%
%%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
%%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
%%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
%\end{proof}
This result is perfectly analogous to what happens in the metric models recalled in the introduction, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
It also entails that for any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$, $\varphi$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.

@@

\end{tikzpicture}
\caption{\small Plot of the tropical polynomials $\varphi_1,\varphi_2,\varphi_3,\varphi_4$ (from top to bottom), and of their limit tLs $\varphi$ (in violet).}
\label{fig:plot1}
%\end{figure} 
\end{wrapfigure} %%

Let us now look at what happens with tLs, i.e.~when considering the full exponential $!$.
First, we consider the interesting case of tLs with \emph{finitely many variables}. 
Let us start by an example, shown in Fig~\ref{fig:plot1}: the 1-variable tLs $\varphi(x)=\inf_{i}\{x+2^{-i}\}$ behaves \emph{locally} like the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{x+2^{-i}\}$. 
%In particular, for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$. More, precisely, $\varphi$ coincides with $\varphi_{n}$ for $\epsilon \geq 2^{-(n+1)}$, the smallest tropical root of $\varphi_{n}$.
However, at
%
%$x\in [\epsilon,\infty]$, with $\epsilon>0$
%It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
 $x=0$ we have that $\varphi(x=0)=\inf_{i\in\N} 2^{-i}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
Also, while the derivative of $\varphi$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
This phenomenon is reminiscent of [Example 7, \cite{Ehrhard2005}].
%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019 which actually motivated our first investigations.
In fact, these properties are shared by all tLs with {finitely} many variables, as shown by the following result (we identify $!\set{1,\dots,k}$ with $\N^k$, so the matrix of a tLs $f$ with variables $x=x_1,\dots,x_k$ can be given as a $\matr f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, where $nx$ is the scalar product).
